---
title: "Reproducibility scripts - supplementary material"
author: "Theodor Balan, Hein Putter"
date: "`r Sys.setenv(LANG = 'en_US.UTF-8'); format(Sys.Date(), '%d %B %Y')`"
output:
  md_document:
    variant: markdown_github
  pdf_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Required packages
```{r}
library(survival)
library(coxme)
library(frailtyEM)
library(tidyverse)
```


# EORTC data

```{r}
# These data unfortunately not available for readers
dat <- read.csv("EORTC_10854.csv", stringsAsFactors = FALSE)

dat$periop <- 
  factor(dat$periop, 
         levels=c("no periop chemo","periop chemo"))
dat$surgery <- 
  factor(dat$surgery, 
         levels=c("mastectomy with RT","mastectomy without RT","breast conserving"))
dat$tusi <- 
  factor(dat$tusi, 
         levels=c("<2 cm","2-5 cm",">5 cm"))
dat$nodal <- 
  factor(dat$nodal, 
         levels=c("node negative","node positive"))
dat$age50 <- 
  factor(dat$age50, 
         levels=c("<=50",">50"))
dat$adjchem <- 
  factor(dat$adjchem, 
         levels=c("no adj chemo","adj chemo"))
dat$tam <- 
  factor(dat$tam, 
         levels=c("no tam","tam"))

dat$hospno <- factor(dat$hospno)

```

## Figure 9: histogram of center sizes
```{r}
dat %>% 
  group_by(hospno) %>% 
  summarize(npat = n()) %>% 
  ggplot(aes(x = npat)) + geom_histogram() + 
  labs(x = "Center size") +
  # scale_x_continuous(breaks = c(0, 50, 100, 182, 304, 602, 880)) +
  scale_x_continuous(breaks = seq(0, 900, by=50)) +
  theme_bw()
```


## Figure 10: Kaplan-Meier survival estimates, overall and by center
```{r}
# data set with Kaplan-Meier curves per center
sf <- survfit(Surv(survyrs, survstat) ~ strata(hospno), dat)
dd <- data.frame(time = sf$time, surv = sf$surv, center = as.factor(rep(1:length(sf$strata), sf$strata))) 
startpoint <- data.frame(center = unique(dd$center), time = 0, surv = 1)
dd <- rbind(dd, startpoint)

# data set with overall Kaplan-Meier curve
sf2 <- survfit(Surv(survyrs, survstat) ~ 1, dat)
dd2 <- data.frame(time = sf2$time, surv = sf2$surv) 

dd %>% 
  ggplot(aes(x = time, y = surv)) + 
  geom_step(aes(group = center),  show.legend = FALSE, alpha = 0.2) + 
  geom_step(data = dd2) + 
  theme_classic() + 
  labs(x = "Years since randomisation", y = "Survival")
```


## Table 1: Estimates from the Cox model, the fixed effects model, the gamma and log-normal frailty models. Estimated center effects of the fixed effects model have been omitted from the table.
```{r}
m_cph <- coxph(Surv(survyrs, survstat) ~ surgery + tusi + nodal + age50 + adjchem +
        tam + periop + cluster(hospno), dat, ties = "breslow")
m_cph

m_fixed <- coxph(Surv(survyrs, survstat) ~ surgery + tusi + nodal + age50 +
                 adjchem + tam + periop + hospno, dat, ties = "breslow")

m_fixed
m_fr_cov <- coxph(Surv(survyrs, survstat) ~ frailty(hospno) + surgery + tusi + nodal + age50 +
                    adjchem + tam + periop, data = dat, ties = "breslow")
m_fr_cov


m_emfrail <- emfrail(Surv(survyrs, survstat) ~ cluster(hospno) + surgery + tusi + nodal + age50 +
                       adjchem + tam + periop,
                     data = dat)
summary(m_emfrail)
# -5450.928 likelihood

m_fr_cov_lnorm <- coxme(Surv(survyrs, survstat) ~ (1|hospno) + surgery + tusi + nodal + age50 +
                          adjchem + tam + periop, data = dat, ties = "breslow")

m_fr_cov_lnorm
# -5449.445 
```

## Figure 11: Center effects from the fixed effects and frailty models, expressed in hazard ratios
```{r}
# Obtain frailty estimates
sm <- summary(m_emfrail)
frailties <- sm$frail

# Obtain center estimates and calculate standard error
centers <- grep("hospno", names(m_fixed$coefficients))
nc <- length(centers)
var_centers <- m_fixed$var[centers, centers]

varmean <- sum(var_centers)/ nc^2

newse <- sqrt(c(varmean, diag(var_centers) + varmean - 2/nc * apply(var_centers, 1, sum)))

# Arrange frailty estimates and fixed effects estimates
logz <- frailties %>% 
  mutate_if(is.numeric, log) %>% 
  rename(estimate = z, ymin = lower_q, ymax = upper_q) %>% 
  mutate(type = "frailty")
  
fe <- data.frame(beta = c(hospnoA = 0, m_fixed$coefficients[centers]), sd = newse) %>% 
  rownames_to_column("Center") %>% 
  mutate(Center = substr(Center, 7, 9)) %>% 
  mutate(sbeta = sum(beta) / 15) %>% 
  mutate(beta = beta - sbeta) %>% 
  arrange(beta) %>%
  mutate(ord = 1:n()) %>% 
  mutate(Center = as.character(Center)) %>% 
  mutate(ymin = beta - 1.96 * sd, ymax = beta + 1.96 * sd) %>% 
  rename(id = Center, estimate = beta) %>% 
  select(-sd, -sbeta, -ord) %>% 
  mutate(type = "fixed effects") %>% 
  mutate(ord = 1:n()) 

ord_fe <- data.frame(id = fe$id, ord = fe$ord)
logz <- logz %>% 
  right_join(ord_fe)

datt <- bind_rows(logz, fe) 

# Finally, the plot
datt %>% 
 mutate_at(vars(estimate, ymin, ymax), .funs = exp) %>% # make it exp()
  ggplot(aes(x = ord, y = estimate)) + 
  geom_point(aes(colour = type), 
             position = position_dodge(.9)) + 
  geom_errorbar(aes(ymin = ymin, ymax = ymax, colour = type, linetype = type),
                position = position_dodge()) + 
  scale_x_continuous(labels = as.character(datt$id)[1:15], 
                     breaks = seq_along(datt$estimate)[1:15]) + 
  scale_y_continuous(trans = "log", 
                     breaks = seq(from = 0.5, to = 2.5, by = .5)) + 
  labs(x = "Center", 
       y="Center effect (hazard ratio)") + 
  theme_bw() + 
  theme(panel.grid.minor = element_blank()) +  
  guides(colour = guide_legend(title = element_blank()), linetype = guide_legend(title = element_blank())) + 
  scale_colour_brewer(palette = "Set1")
```

## Extra: Checking proportional hazards
```{r}
# checking proportional hazards assumption

# in Cox model
cox.zph(m_cph)

# in Cox model with log-frailties as offset
logz_long <- log(m_emfrail$frail)[dat$hospno]

m_cph_offset_gamma <- coxph(Surv(survyrs, survstat) ~ surgery + tusi + nodal + age50 + adjchem +
                 tam + periop + offset(logz_long), dat, ties = "breslow")
cox.zph(m_cph_offset_gamma)

```



# CGD data

## Load data
```{r}
data(cgd)
# head(cgd)

# rebrand id
cgd <- cgd %>% 
  mutate(id = as.numeric(as.factor(id))) 
```

## Figure 12: Event history of the CGD data. The length of the line indicates the length of follow-up, and the dots indicate the infections
```{r}
cgd %>% 
  ggplot() +
  geom_segment(aes(x = tstart, xend = tstop, y = id, yend = id)) + 
  geom_point(data = filter(cgd, status == 1), aes(x = tstop, y = id)) + 
  theme_bw() + 
  labs(x = "time", y = "patient")


```

## Figure 13: Histogram of number of events per individual

```{r}
cgd %>% 
  group_by(id) %>% 
  summarize(n_events = sum(status)) %>% 
  ggplot(aes(x = n_events)) + 
  geom_bar() + 
  labs(x= "Number of events / individual") + 
  theme_bw()
```


## Extra: Grid search for optimal m for Compound Poisson
```{r}
mvals <- seq(from = 0.1, to = 2, by = 0.2)
models <- lapply(mvals, function(x) emfrail(formula = Surv(tstart, tstop, status) ~ sex +
    treat + age + propylac + inherit + steroids + cluster(id), data = cgd, 
    distribution = emfrail_dist(dist = "pvf", pvfm = x)) )

likelihoods <- sapply(models, function(x) x$loglik[2])

mvals[which.max(likelihoods)]

```

## Table 2: Estimates of the regression coefficients and fit summary from the Cox model and shared frailty models, with gamma, inverse Gaussian, positive stable and compound Poisson (with parameters 0.5 and 1.1) distributions, fitted on the CGD data

```{r}
m_cph <- coxph(Surv(tstart, tstop, status) ~ treat + age +inherit + steroids, cgd)
# summary(m_cph)


m_gamma_emf <- emfrail(Surv(tstart, tstop, status) ~ treat + sex + age +inherit + steroids + cluster(id), cgd)
# summary(m_gamma_emf)

mod_ig <- emfrail(formula = Surv(tstart, tstop, status) ~  treat + sex +
                    age + inherit + steroids + cluster(id), 
                  distribution = emfrail_dist(dist = "pvf"),
                  data = cgd)

mod_cp <- emfrail(formula = Surv(tstart, tstop, status) ~  treat + sex +
                    age + inherit + steroids + cluster(id), 
                  distribution = emfrail_dist(dist = "pvf", pvfm = 0.5),
                  data = cgd)

mod_cp_11 <- emfrail(formula = Surv(tstart, tstop, status) ~  treat + sex +
                    age + inherit + steroids + cluster(id), 
                  distribution = emfrail_dist(dist = "pvf", pvfm = 1.1),
                  data = cgd)

mod_stab <- emfrail(formula = Surv(tstart, tstop, status) ~ treat + sex + 
                      age + inherit + steroids + cluster(id), 
                    distribution = emfrail_dist(dist = "stable"),
                    data = cgd)

```

## Figure 14: Histogram of estimated frailties

```{r}
autoplot(m_gamma_emf, type = "hist") + 
  theme_bw()
```

# R session Info
```{r}
sessionInfo()
```

